What We Eval in the Shadows

### 2021-04-17 paper/main.tex
Proofreading Section 6

In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them on concrete instances.

@@

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function in
    the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases, the variable is a
    symbol and so the \eval seems superfluous. The \eval is however necessary so
    one can specify parent classes using \c{pkg\_name::class\_name}, as the
    \c{::} operator is a function call in R.

@@

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions capturing
    any assignments that happen in it, as samples of Markov chains.

@@

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for a later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation with statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
    hazards regression model.} contains:
    %
    \begin{lstlisting}[gobble=2]
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
    fragment of a program to be evaluated in a special way and in a special
    environment. The latter is what makes it different from 0-argument closure.

@@

% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd

% \subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
% \c{match.call} walks up the call stack, captures the code that invoked the
% currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments. The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified. This pattern is recognizable
% by the fact that the expression is a call and the target environment is that of
% the parent.

% % We have quantitative data on that if needed...
% This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}. The R documentation of \c{match.call} explicitly describes this usage.

% In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
% is evaluated in the same environment the original call was.

% \begin{lstlisting}
% Call <- match.call()
% tform <- Call[c(1,indx)]  # only keep the arguments we wanted
% tform[[1L]] <- quote(stats::model.frame)  # change the function called
% mf <- eval(tform, parent.frame())
% \end{lstlisting}

% A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
% In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data. Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}. The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
% \begin{lstlisting}
% computeMDS95 <- function (data, Vegetables, ...) {
% 	arguments <- as.list( match.call())
% 	Vegetables <- eval(arguments$Vegetables, data)
% \end{lstlisting}

% \c{substitute} can be used to capture only a part of the call, \ie one of the arguments. For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
% \begin{lstlisting}
% xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
% subset <- eval(substitute(subset), data, environment(x))
% \end{lstlisting}

% In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}. This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.

% \subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.

%  A call to \c{library(packageName)}, which loads libraries in R, for instance, will
% populate environment \c{package:packageName} using \eval. It is a \emph{base} function
% though, so we do not see it in the CRAN dataset. However, R still makes it
% possible to access any package environments, which we see in
% \packageNbPackageEnvPackages different packages in our corpus. Those packages perform
% introspection, to represent and search any environment by name, in package
% \emph{envnames}, to containerize tests, as with \emph{testthat} and
% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
% implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.

% \subsubsection{Code Generation.} The more traditional use of \eval is to execute code
% that was assembled by the programmer into a string. Here we show the method
% \c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
% explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
% \begin{lstlisting}
% eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
% \end{lstlisting}

% In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}. In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.

% \begin{lstlisting}
% p <- lapply(paste0('PC', 1:9), function(pc){
% 	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
% })
% \end{lstlisting}

% The \parse and \eval combination is
% used by the \source function in the base library to load R code from a file
% in the current workspace.

%  Code generation is often used as a means to reduce boilerplate code;
% simple and repetitive code can easily be replaced with judicious use of \eval.
% For example, the \c{data.table} package uses \eval to calls the \c{options}
% function with named arguments taken from a vector of strings. While the benefits
% are limited in this example, it is an attractive tool for programmers.
% \begin{lstlisting}
%   opts = c("datatable.verbose"="FALSE", # ...many others
%   for (i in names(opts))
%     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
% \end{lstlisting}
% This pattern is a special case of code generation, recognizable by the
% fact that \eval is executed in a loop.

% Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions. On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline. In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse} on strings with \eval.

% \todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp. But at the same type, we lack here an example of code generation using quite, enquote or substitute.}

% \todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. For example,
the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
uses \eval to resolve a string into function and another to call it and assign
its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
or even to a one-liner \c{do.call(paste0("q",d), as.list(dp, ...))}.

@@

\Eval in R is mostly used for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as
function or method call, or assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\c{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \c{get}, and assignment in any environment,
with \c{assign}. Building a call and executing can be done with \c{do.call}.
Nevertheless the many of the observed \eval usage in R, in particular the one
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a mean to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically, by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example, the it checks that packages
do not call \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package name space.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

>>>>

In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
expressions passed to \eval. Here, we illustrate them in concrete instances.

@@

    Another example is the most often called \eval site in CRAN:
    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function in
    the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
    find the parent class. In the vast majority of cases, the variable is a
    symbol, so the \eval seems superfluous. The \eval is however necessary so
    one can specify parent classes using \c{pkg\_name::class\_name}, as the
    \c{::} operator is a function call in R.

@@

    There are also more sophisticated cases. The \c{plyr::allocate\_column}
    function uses \eval for a deferred assignment to fill missing values in a
    column. Package \c{overture} repeatedly evaluates R expressions capturing
    any assignments that happen in it as samples of Markov chains.

@@

    Call reflection refers to the combination of \eval with \c{match.call}, a
    function that returns the current call with all its arguments captured as
    expressions. This is used mostly in the following circumstances: (1) to
    record a call for later use in a possibly different environment, (2) pass
    most of the captured call arguments to another function, or (3) evaluate
    arguments of the function in different environments. This is widely used in
    relation to statistical modeling in combination with a model fitting
    function or with \c{model.frame} and related functions. For example the
    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
    hazards regression model.} contains:
    %
    \begin{lstlisting}[gobble=2]
    Call <- match.call()
    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
    tform[[1L]] <- quote(stats::model.frame)  # change the function called
    mf <- eval(tform, parent.frame())
    \end{lstlisting}

  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
    the case that the block is only passed to \eval rather than being directly
    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
    fragment of a program to be evaluated in a particular way and in a particular
    environment. The latter is what makes it different from 0-argument closure.

@@

@@

Similarly to JavaScript, there are also unnecessary uses of \eval. For example,
the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
uses \eval to resolve a string into function and another to call it and assign
its results into a variable:
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
  }
\end{lstlisting}
  In both cases, there is no need for \eval and the function can be rewritten as
\begin{lstlisting}
  function (R, d="norm", dp, ...) {
	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
  }
\end{lstlisting}
or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.

The problem is that these cases are not easy to spot. Because of the
limits of dynamic analysis, we only have a partial coverage and cannot determine
if for example, \eval sites classified as \c{X, V} will not be eventually called
with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.

@@

\Eval in R is used chiefly for meta-programming purposes and to access
environments other than the local one. In Javascript, patterns are ad-hoc, such
as json loading and parsing, and can often be rewritten without \eval, such as
function or method call, or assignment to a local variable or an object
property. For R, the data suggests that in some cases, the use of \eval in the
\c{X, V, \$, <-, FUN} and the function calls related to code generation could
replaced by more specific functions. For instance, variable lookup in any
environment can be performed with \c{get}, and assignment in any environment,
with \c{assign}. Building a call and executing can be done with \c{do.call}.
Nevertheless, many of the observed \eval usage in R, particularly, the ones
related to statistical modeling and DSLs are more sophisticated than those
reported for JavaScript.

Finally, let us mention an anecdotal use of \eval as a means to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example,  it checks that packages
do not call \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package namespaces.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.
